##IMPACT OF WEATHER EVENTS ON HEALTH AND ECONOMY



###Synopsis

Storms and severe weather events in the U.S. can negatively impact both public health and the 
economy, in terms of fatalities, injuries and property damage. This project will analyze data
from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and address
which type of events are the most harmful to population health and to the economy. 

An analysis of the storm data and the resulting plots show that **TORNADO**, **EXCESSIVE HEAT** and **FLOOD**
are the top 3 weather events that cause the most harm to the population in terms of total fatalities and injuries,
across the United States from 1989 to 2011.

Meanwhile, **FLOOD**, **HURRICANE/TYPHOON** and **STORM SURGE** are the top 3 weather events that cause the most 
economic damage to crops and property across the United States from 1989 to 2011.

The number of weather events breached the 10,000 mark starting 1989.  Hence this report analyzed a subset of the 
original data with those only from 1989 to 2011.  However a separate analysis using the entire storm data from
1950 to 2011 would have shown similar results to those from 1989 to 2011



###Data

The storm data is available in csv format from the course website: 

- [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

Further documentation of the database is available at:

- National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)



###Retrieving data


Load the storm data into a data frame called data:

```{r cache=TRUE}
URL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "repdata-data-StormData.csv"
download.file(URL, destfile)
data <- read.csv("repdata-data-StormData.csv")
rows <- nrow(data)
cols <- ncol(data)
```

The data frame contains `r rows` rows (observations) and `r cols` columns (variables). 

Quick look at the first few data observations
```{r}
head(data)
```


###Data Processing

Take a subset of the original data which include variables that are relevant to the analysis.
In particular, we will include the date, weather event, health and economic impacts
```{r}
mycol <- c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
mydata <- data[,mycol]
library(data.table)
mydata = data.table(mydata)
dim(mydata)
head(mydata)
```

The events in the database start in the year 1950 and end in November 2011. In the 
earlier years of the database, there are generally fewer events recorded, most likely
due to a lack of good records.  More recent years should be considered more complete.
This code chunk creates a new column called year, strips date and time and counts the 
number of weather events per year. 

```{r}
mydata$year = as.numeric(format(as.Date(mydata$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), 
                                "%Y"))
stormsByYear = mydata[, lapply(.SD, length), by = year, .SDcols = 1]
setnames(stormsByYear, 2, "count")
stormsByYear
```

The **stormsByYear** shows that the number of weather events breached the 10,000 mark starting 1989.
Hence we can use a subset of the original data with those only from 1989 to 2011.  However a separate
analysis using the entire storm data from 1950 to 2011 would show similar results to those from 1989 to 2011. 

```{r}
storm = mydata[year >= 1989]
dim(storm)
```

For economic impact, we shall include property damage and crop damage data.
Transform the data into comparable numeric units as described in the code book.
For example, "B" = billions, "M" = millions, "K" = thousands, "H" = hundreds.

```{r results='hide'}
storm$PROPDMGEXP = as.character(storm$PROPDMGEXP)
storm$PROPDMGEXP[toupper(storm$PROPDMGEXP) == "B"] = "9"
storm$PROPDMGEXP[toupper(storm$PROPDMGEXP) == "M"] = "6"
storm$PROPDMGEXP[toupper(storm$PROPDMGEXP) == "K"] = "3"
storm$PROPDMGEXP[toupper(storm$PROPDMGEXP) == "H"] = "2"
storm$PROPDMGEXP = as.numeric(storm$PROPDMGEXP)
storm$PROPDMGEXP[is.na(storm$PROPDMGEXP)] = 0
storm$PropertyDamage = storm$PROPDMG * 10^storm$PROPDMGEXP
summary(storm$PropertyDamage)

storm$CROPDMGEXP = as.character(storm$CROPDMGEXP)
storm$CROPDMGEXP[toupper(storm$CROPDMGEXP) == "B"] = "9"
storm$CROPDMGEXP[toupper(storm$CROPDMGEXP) == "M"] = "6"
storm$CROPDMGEXP[toupper(storm$CROPDMGEXP) == "K"] = "3"
storm$CROPDMGEXP[toupper(storm$CROPDMGEXP) == "H"] = "2"
storm$CROPDMGEXP[toupper(storm$CROPDMGEXP) == ""] = "0"
storm$CROPDMGEXP = as.numeric(storm$CROPDMGEXP)
storm$CROPDMGEXP[is.na(storm$CROPDMGEXP)] = 0
storm$CropDamage = storm$CROPDMG * 10^storm$CROPDMGEXP
```

Assume that total damage on economy is the sum of property and crop damage

```{r}
storm$TotalDamage = storm$PropertyDamage + storm$CropDamage
```

Assume total health impact is the sum of fatalities and injuries

```{r}
storm$TotalImpact = storm$FATALITIES + storm$INJURIES
```

Aggregate the data by event

```{r}
fatal <- aggregate(FATALITIES ~ EVTYPE, data = storm, FUN = sum)
injury <- aggregate(INJURIES ~ EVTYPE, data = storm, FUN = sum)
fatal_injury <- aggregate(TotalImpact ~ EVTYPE, data = storm, FUN = sum)
propdmg <- aggregate(PropertyDamage ~ EVTYPE, data = storm, FUN = sum)
cropdmg <- aggregate(CropDamage ~ EVTYPE, data = storm, FUN = sum)
totaldmg <- aggregate(TotalDamage ~ EVTYPE, data = storm, FUN = sum)
```


###Results

Get top 10 weather events with highest total health impact and plot the results.

```{r}
fatal_injury10 <- fatal_injury[order(-fatal_injury$TotalImpact), ][10:1, ]
par(mfrow = c(1, 1), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(fatal_injury10$TotalImpact, las = 3, names.arg = fatal_injury10$EVTYPE, 
        main = "Top 10 Weather Events With Highest Total Fatalities and Injuries", 
        ylab = "number of total injuries/fatalities", col = "orange")
```

Get top 10 weather events with highest total damage to crops and property and plot the results.

```{r}
totaldmg10 <- totaldmg[order(-totaldmg$TotalDamage), ][10:1, ]
par(mfrow = c(1, 1), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(totaldmg10$TotalDamage/10^9, las = 3, names.arg = totaldmg10$EVTYPE, 
        main = "Top 10 Weather Events With Highest Total Damages to Crops and Property ", 
        ylab = "Cost of damages ($ billions)", 
        col = "orange")
```


###Conclusion


An analysis of the storm data and the resulting plots show that **TORNADO**, **EXCESSIVE HEAT** and **FLOOD**
are the top 3 weather events that cause the most harm to the population in terms of total fatalities and injuries,
across the United States from 1989 to 2011.


Meanwhile, **FLOOD**, **HURRICANE/TYPHOON** and **STORM SURGE** are the top 3 weather events that cause the most 
economic damage to crops and property across the United States from 1989 to 2011.

The number of weather events breached the 10,000 mark starting 1989.  Hence this report analyzed a subset of the 
original data with those only from 1989 to 2011.  However a separate analysis using the entire storm data from
1950 to 2011 would have shown similar results to those from 1989 to 2011.
